# Powhatan County {#part-4-9}

This chapter is a summary of the major changes to the Powhatan County's population and housing market in the past five years.

```{r setup}

library(tidyverse)
library(tidycensus)
library(sf)
library(sp)
library(plotly)
library(simplevis)
library(tidytext)
library(scales)
library(readxl)
library(zoo)
library(lubridate)
library(janitor)
library(tigris)
library(FinCal)
library(kableExtra)
library(ggiraph)
library(ggtext)
library(hdatools)

years <- 2016:2020
fips <- 51145
name <- "Powhatan County"
costar_name <- "Powhatan County"



```

## Takeaways

- Powhatan County is growing like many other parts of the region --- passing the 30,000 mark in 2020.
- More and more homeowners without children have been moving into the county --- nearly 600 households between 2016 and 2020.
- Unlike other localities, the county saw a decrease of nearly 100 seniors living alone between 2016 and 2020.
- Median renter household income in the county saw a 20 percent increase in recent years, while fewer and fewer renters in the county.
- Although only 29 enrolled students experienced homelessness during the 2019-2020 school year, this was a doubling from 2016-2017 when the count was at 15.

## Demographic and socioeconomic changes

### Population changes

Despite a dip in population from 2014 to 2015, Powhatan County has been on an upward trend for the most part. In 2020, the county reached a population of 30,333 --- an overall increase of 8 percent since 2010.

```{r}
#| label: fig-pop
#| fig-cap: !expr "paste0(name, \": Total Population\")"

# Add choice of population stat (net change, components, etc)

pop <- read_rds("data/rr_pop_data.rds") |> 
  mutate(GEOID = parse_number(GEOID)) |> 
  filter(counttype != "Forecast",
         GEOID == fips)

pop_bar <- ggplot(pop,
                      aes(x = year,
                          y = value,
                          fill = value,
                          data_id = value,
                          tooltip = label_comma()(value))) +
  geom_col() +
  geom_col_interactive(size = 2) +
  scale_fill_gradientn(colours = c("#CAE3C2", "#74BA91", "#519B8D", "#2B4258")) +
  theme_pha() +
  labs(title = paste0(name, ": Total Population"),
       subtitle = "2010 to 2020",
       caption = "**Source:** U.S. Census Bureau Decennial Census and American Community Survey.") +
  scale_y_continuous(labels = label_comma()) +
  theme(axis.title = element_blank(),
        legend.position = "none",
        panel.background = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major.y = element_line(color = "grey95",
                                          size = 0.05)) +
  theme(plot.title.position = "plot",
        plot.caption.position = "plot")

if (knitr::is_html_output()) {
  
  girafe(ggobj = pop_bar,
         height_svg = 4)
} else {
  
  pop_bar
}


```

Like many other rural counties in the region, Powhatan County's population growth in recent years has been due to domestic migration. Between 2020 and 2021, 772 new residents came to the county from somewhere else within the country. 

```{r}
#| label: fig-comp
#| fig-cap: !expr "paste0(name, \": Components of population change\")"

change <- read_rds("data/rr_comp_change.rds") |> 
  filter(year %in% as.character(years) | year == "2021") |> 
  filter(NAMELSAD == name)

pop_change <- ggplot(change,
       aes(x = year,
           y = value,
           fill = component,
           data_id = value,
           tooltip = value)) +
  geom_col() +
  geom_col_interactive(size = 2) +
  facet_wrap(~component) +
  theme_pha() +
  scale_fill_pha() +
  labs(title = paste0(name,": Components of population change"), 
       subtitle = "2016 through 2021",
       caption = "**Note:** 2020 components of change data not available.\n**Source:** U.S. Census Bureau, Population Estimates Program.")  +
  scale_y_continuous(labels = label_number(big.mark = ",", style_positive = "plus", style_negative = "minus")) 


if (knitr::is_html_output()) {
  
  girafe(ggobj = pop_change,
         height_svg = 4)
  
} else {
  
  pop_change
  
  }

```

### Household characteristics

The county has seen a decreasing number of renter households across the board. But homeowners continue to come to the county --- especially homeowner households with no children. By 2020, there was an estimated 590 more homeowners with no children than in 2016.

```{r}
#| label: fig-children
#| fig-cap: !expr "paste0(name, \": Change in households with children by tenure\")"

# Add choice of household stat (HHs with children, HHs with seniors, subfamilies, etc)

b25115 <- read_rds("data/b25115_data.rds") |> 
  filter(NAME == name)

ch_plot <- ggplot(b25115,
       aes(x = children, 
           y = change, 
           fill = tenure,
           data_id = change,
           tooltip = label_comma()(change))) +
  geom_col(position = "dodge") + 
  geom_col_interactive(position = "dodge", size = 2) +
  theme_pha() + 
  scale_fill_pha() +
  facet_wrap(~children, nrow = 1, scales = "free_x") +
  scale_y_continuous(labels = label_number(style_positive = "plus", big.mark = ",")) +
  labs(title = paste0(name, ": Change in households with children by tenure"),
       subtitle = "2016 to 2020",
       caption = "**Source:** U.S. Census Bureau, American Community Survey, Table B25115.") 


if (knitr::is_html_output()) {
  
  girafe(ggobj = ch_plot,
         height_svg = 4)
  
} else {
  
  ch_plot
  
  }

```
The greatest increases in the senior population in the county have been among seniors in family households (either living with a spouse or are the head of household). In contrast to other localities, Powhatan County is seeing a decline in seniors in nonfamily households.


```{r}
#| label: fig-seniors
#| fig-cap: !expr "paste0(name, \": Change in senior population by living arrangement\")"

b09020 <- read_rds("data/b09020_data.rds") |> 
  filter(NAME == name) |> 
  mutate(family = str_replace(family, "Group quarters", " "),
         family = fct_relevel(family, "Family", "Nonfamily", " "))

snr_plot <- ggplot(b09020,
       aes(y = reorder_within(relationship, change, family),
           x = change,
           fill = family,
           data_id = change,
           tooltip = change)) +
  geom_col(position = "dodge") +
  geom_col_interactive(position = "dodge", size = 2) +
  theme_pha() +
  scale_fill_pha() +
  scale_y_reordered() +
  scale_x_continuous(labels = label_comma()) +
  facet_grid(rows = vars(family), scales = "free_y", space = "free", switch = "y") +
  #scale_y_discrete(labels = function(x) str_wrap(x, width = 10)) +
  labs(title = paste0(name, ": Change in senior population by living arrangement"),
       subtitle="2016 to 2020",
       fill="Tenure",
       caption = "**Source:** U.S. Census Bureau, American Community Survey, Table B09020.")  +
  theme(strip.placement = "outside",
        panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_line(color = "#e2e4e3",
                                          size = 0.05))

if (knitr::is_html_output()) {
  
  girafe(ggobj = snr_plot,
         height_svg = 4)
  
} else {
  
  snr_plot
  
  }

```

### Income and wages

Both homeowner and renter median household income has been on the incline from 2016 to 2020 in the county. For homeowners, there has been an increase of 8 percent, while renters experienced a 20 percent increase.

```{r}
#| label: fig-income
#| fig-cap: !expr "paste0(name, \": Median houshold income by tenure\")"

# Add choice of income stat (probably median household income of some kind)

b25119_cpi <-read_rds("data/b25119_cpi.rds") |> 
  filter(NAME == name)

inc_plot <- ggplot(b25119_cpi, 
       aes(x = year,
           y = dollars20,
           fill = tenure,
           data_id = dollars20,
           tooltip = dollar_format()(dollars20))) +
  geom_col() +
  geom_col_interactive(size =2) +
  theme_pha() +
  scale_fill_pha() +
  facet_wrap(~tenure) +
  scale_y_continuous(labels = dollar_format(), limits = c(0, NA)) +
  labs(title = paste0(name, ": Median houshold income by tenure"), 
       color = "Tenure",
       subtitle = "2016 to 2020 | Adjusted to 2020 dollars",
       caption = "**Source:** U.S. Census Bureau, American Community Survey")

if (knitr::is_html_output()) {
  
  girafe(ggobj = inc_plot,
         height_svg = 4)
  
} else {
  
  inc_plot
  
  }


```

### Persons with disabilities

In the county, there has been an increase in individuals with independent living difficulties that are 65 years old and older (+204) between 2016 and 2020. For individuals younger than 65, there has been a decline in individuals with independent living difficulties.

```{r}
#| label: fig-ind-living
#| fig-cap: !expr "paste0(name, \": Net change in individuals with independent living difficulties by age\")"

# Add disability stat (probably independent living difficulty)

b18107 <- read_rds("data/b18107_data.rds") |> 
  filter(NAME == name) |> 
  group_by(age) |> 
  summarise(change = sum(change))

ind_plot <- ggplot(b18107,
       aes(x = age,
           y = change,
           fill = age,
           data_id = change,
           tooltip = change)) +
  geom_col() +
  geom_col_interactive(size = 2) +
  theme_pha() +
  scale_fill_pha() +
  labs(title=paste0(name, ": Net change in individuals with\nindependent living difficulties by age"),
       subtitle="2016 to 2020",
       caption="**Source:** U.S. Census Bureau, American Community Survey, Table B18107.") +
  scale_y_continuous(labels = label_comma()) 

if (knitr::is_html_output()) {
  
  girafe(ggobj = ind_plot,
         height_svg = 4)
  
} else {
  
  ind_plot
  
  }

```



## Housing supply and market changes

### Homeownership

Powhatan County median home price has been on the rise. From a low of \$283,507 in March 2017 to a high of \$522,181 in May 2022, median home price in the county has increased nearly 84 percent. Although the over \$500,000 median home price may represent a seasonal spike, median home price has generally been trending upward and likely to stay above \$500,000 in the coming years.

```{r}
#| label: fig-sales
#| fig-cap: !expr "paste0(name, \": Median home sales price\")"

# Median sales price over time

price <- read_csv("data/powhatan_med_sales.csv") |> 
  clean_names() |> 
  mutate(geography = name) |> 
  pivot_longer(cols = starts_with("sale"),
    names_to = "label",
    values_to = "med_price",
    values_drop_na = TRUE) |> 
  mutate(year = substr(label, 19,23)) |> 
  select(geography, year, month, med_price)

price$month <- match(price$month, month.abb)

price <- price |> 
  mutate(date = make_date(year, month)) |> 
  mutate(med_price = parse_number(med_price))

cpi <- read_csv("data/CPIAUCSL.csv") |> 
  select(date = DATE, cpi = CPIAUCSL)

price_adj <- price |> 
  left_join(cpi, by = "date") |> 
  mutate(dollars = ((296.761/cpi)*med_price)) |> 
  select(date, med_price, dollars) |> 
  pivot_longer(2:3,
               names_to = "type",
               values_to = "dollars") |> 
  filter(type == "dollars")

sales_plot <- ggplot(price_adj,
       aes(x = date,
           y = dollars,
           color = dollars,
           group = 1)) +
  geom_line(size = 1) +
  geom_point_interactive(
    aes(data_id = dollars,
        tooltip = label_dollar()(dollars)),
    size = 0.5) + 
  theme_pha() +
  scale_color_gradientn(colours = c("#CAE3C2", "#74BA91", "#519B8D", "#2B4258")) +
  scale_y_continuous(labels = label_dollar()) +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  labs(title = paste0(name, ": Median home sales price"),
       subtitle = "January 2017 to July 2022",
       caption = "**Source:** Central Virginia Regional Multiple Listing Service.") +
  geom_smooth(method = "lm", color = "#Be451c")

if (knitr::is_html_output()) {
  
  girafe(ggobj = sales_plot,
         height_svg = 4)
  
} else {
  
  sales_plot
  
  }

```

### Rental

Renter households have been in decline since 2016 when roughly one in ten households in the county were renters. This share has dropped by four percentage points down to 7 percent renter household in 2020. Despite the declining number of renter households in the county, 2021 saw the completion of 200 new rental units at the eastern edge of the county.

```{r}
#| label: fig-renters
#| fig-cap: "Renter households in Powhatan County"

b25003_vars <- load_variables(2020, "acs5") |> 
  filter(str_sub(name, end = 7) %in% "B25003_")

b25003_raw <- map_dfr(years, function(yr){
   b25003_pull <- get_acs(
    geography = "county",
    state = "VA",
    table = "B25003",
    year = yr,
    survey = "acs5"
  ) |> 
    mutate(year = yr)
})

b25003_vars_cleaned <- b25003_vars |> 
  separate(label, into = c("est", "total", "tenure"), sep = "!!") |> 
  select(variable = name, tenure) |> 
  mutate(tenure = case_when(
    tenure == "Owner occupied" ~ "Homeowner",
    tenure == "Renter occupied" ~ "Renter"
  )) |> 
  drop_na()

b25003_raw <- b25003_raw |> 
  filter(GEOID %in% fips)

b25003_data <- b25003_raw |> 
  right_join(b25003_vars_cleaned, by = "variable") |> 
  select(NAME, GEOID, year, tenure, estimate) |> 
  mutate(NAME = str_remove_all(NAME, ", Virginia")) |>
  group_by(NAME, year) |> 
  mutate(pct = (estimate/sum(estimate))) |> 
  filter(tenure == "Renter")

renter_plot <- ggplot(b25003_data,
       aes(x = year,
           y = pct,
           fill = pct,
           data_id = pct,
           tooltip = label_percent()(pct))) + 
  geom_col() +
  geom_col_interactive(size = 2) +
  theme_pha() +
  scale_fill_gradientn(colours = c("#CAE3C2", "#74BA91", "#519B8D", "#2B4258")) +
  labs(title = "Percent of renter households in Powhatan County", 
       fill = "Percent of total households",
       subtitle = "From 2016 to 2020",
       caption = "**Source:** U.S Census Bureau. American Community Survey 2016-2020 5-Year Estimates. Table B25003.") +
  scale_y_continuous(labels = percent_format())


if (knitr::is_html_output()) {
  
  girafe(ggobj = renter_plot,
         height_svg = 4)
  
} else {
  
  renter_plot
  
  }


```
CoStar lists three multifamily properties in the county; two are located along Old Buckingham Road near the county seat, while the third is located at the eastern edge of the county near Midlothian. There was a 39 year period between the construction of both Powhatan Apartment buildings and the Artistry. 

The development of the Artistry at Winterfield potentially represents growing housing demand from the western edge of Chesterfield County's Midlothian community. This luxury multifamily property has an average asking rent of \$1,888. 

```{r}
#| label: tbl-mf-list
#| tbl-cap: "Powhatan County: Multifamily properties"

names <- c("Artistry at Winterfield", "Powhatan Apartments", "Powhatan Apartments")
units <- c(200, 11, 11)
rating <- c(4, 2, 3)
year <- c(2021, 1982, 1982)
vacancy <- c("7.8%", "4.6%", "4.6%")

mf_tbl <- data.frame(names, year, rating, vacancy, units)

mf_tbl |> kable(align = "lcccr",
        col.names = c("Property name", "Year built", "CoStar rating", "Vacancy rate", "Units")) |> 
  kableExtra::kable_styling(
    bootstrap_options = c("condensed", "striped", "hover"))


```

### Housing assistance

The National Housing Preservation Database lists no federally-assisted housing properties in New Kent County. 

### Naturally-occurring affordable housing

Powhatan Apartments (both locations) represent the county's naturally-occurring affordable housing based on the definition of NOAH outlined in this report. Rent data is unavailable for Powhatan Apartments via CoStar.

In addition to Powhatan Apartments, there is a small manufactured home community located near the cross roads of Maidens Road (US-522) and Huguenot Trail that consists of at least six homes.

```{r}
#| label: fig-mhc
#| fig-cap: !expr "paste0(name, \": Manufactured home communities\")"

mhc <- st_read("https://services1.arcgis.com/Hp6G80Pky0om7QvQ/arcgis/rest/services/Mobile_Home_Parks/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson", quiet = TRUE) |> 
  filter(STATE == "VA",
         COUNTY == "POWHATAN")

library(leaflet)

leaflet(mhc) |> 
  addTiles() |> 
  addCircleMarkers()

```

## Gap analysis

### Affordability of current housing stock

With increasing renter household incomes in the county, the gap between income needed to afford the typical home price and typical renter income has been narrowing. In 2016, the gap stood at nearly \$12,000, but by 2020 the gap had decreased to just below \$7,000. 

```{r}
#| label: fig-income-own-gap
#| fig-cap: !expr "paste0(name, \": Income needed to afford median home price\nversus median renter income\")"

# Compare local renter income versus local home prices

rr_sales_annual <- read_csv("data/rr_annual_sales.csv") |> 
  pivot_longer(cols = 2:9,
               names_to = "locality",
               values_to = "price") |> 
  clean_names() |> 
  filter(locality == costar_name)

int_annual <- read_csv("data/fredmac_int_annual.csv")

downpayment <- 0.05 # 5% downpayment
closingcosts <- 0.015 # 1.5% closing costs
utilities <- 250 # Assume $250/month for utilities

b25119_cpi <- b25119_cpi |> 
  filter(tenure == "Renter")

sale <- rr_sales_annual |> 
  left_join(b25119_cpi, by = "year") |> 
  left_join(int_annual, by = "year") |> 
  drop_na() |> 
  mutate(principal = price - (price*downpayment)) |> 
  mutate(loanamt = principal/(1-closingcosts)) |> 
  mutate(mortgage = abs(pmt((annual_int/12), 360, loanamt, 0)) + 250) |> 
  mutate(inc_needed = ((mortgage*10)/2.8)*12) |> 
  select(year, inc_needed, cdollars)

slgap <- sale |> 
  pivot_longer(2:3,
               names_to = "data",
               values_to = "value") |> 
  mutate(data = case_when(
    data == "cdollars" ~ "Renter median household income",
    data == "inc_needed" ~ "Income needed to afford median home price"
  ))

slgap_plot <- ggplot(slgap,
       aes(x = year,
           y = value,
           fill = str_wrap(data, 25),
           data_id = value,
           tooltip = dollar_format()(value))) +
  geom_col(position = "dodge") +
  geom_col_interactive(position = "dodge", size = 2) +
  scale_y_continuous(labels = label_dollar()) + 
  theme_pha() +
  scale_fill_pha() +
  labs(title = paste0(name, ": Income needed to afford median home price\nversus median renter income"),
       subtitle = "2016 to 2020",
       fill = "Nominal income",
       caption = "**Sources:** U.S. Census Bureau, American Community Survey, Table B25119\nand Central Virginia Regional Multiple Listing Service.") +
  theme(legend.position = "top")

if (knitr::is_html_output()) {
  
  girafe(ggobj = slgap_plot,
         height_svg = 4) 
  
} else {
  
  slgap_plot
  
  }


```

As of 2018, there was a 70 unit rental home shortage for households making less than 80 percent AMI. This was a major decline from 2015 when the shortage was 215 --- mainly for 30 percent AMI or less households. The deficit has decreased significantly among extremely and very low-income households, but has increased for higher income households.

```{r}
#| label: fig-gap
#| fig-cap: !expr "paste0(name, \": Rental housing gap by AMI\")"

# Add choice of affordability stat (surplus/deficit, income vs prices, etc)

gap <- read_rds("data/tb_18_match.rds") |> 
  filter(county == name) |> 
  group_by(year, household_income, gapcode) |>
  summarise(estimate = sum(estimate)) |>
  mutate(estimate = case_when(
    gapcode == "Gap" ~ estimate * -1,
    gapcode == "Matches or less than income" ~ estimate
  )) |>
  mutate(gapcode = case_when(
    gapcode == "Gap" ~ "Shortage of units",
    gapcode == "Matches or less than income" ~ "Households at or less than income"
  )) |>
  filter(household_income != "80 percent AMI or greater")

gap_plot <- ggplot(gap,
       aes(x = year, 
           y = estimate,
           fill = gapcode,
           data_id = estimate,
           tooltip = estimate)) +
  geom_col() + 
  geom_col_interactive(size = 2) + 
  facet_grid(~household_income) + 
  theme_pha() +
  scale_fill_pha() + 
  scale_y_continuous(labels = label_comma()) +
  labs(title = paste0(name, ": Rental housing gap by AMI"),
       subtitle = "2015 to 2018",
       caption = "**Source:** U.S. Department of Housing and Urban Development,\nComprehensive Housing Affordability Strategy (CHAS), Table 7.") +
  theme(legend.position = "top")

if (knitr::is_html_output()) {
  
  girafe(ggobj = gap_plot,
         height_svg = 4) 
  
} else {
  
  gap_plot
  
  }

```

### Impact of housing costs

The share of cost-burdened renters has decreased from nearly 38 percent of renter households to 26 percent in a short period. For homeowners, the share of cost-burdened households has decreased by 8 percentage points --- going from 24 percent to 16 percent over four years.

```{r}
#| label: fig-cb
#| fig-cap: !expr "paste0(name, \": Share of cost-burdened households by tenure\")"

cb <- read_rds("data/cb_7.rds") |> 
  filter(county == name) |> 
  group_by(year, tenure, cb_group) |> 
  summarise(estimate = sum(estimate)) |> 
  mutate(pct = estimate/sum(estimate))

cb_plot <- ggplot(cb,
       aes(x = year, 
           y = pct, 
           fill = cb_group,
           data_id = pct,
           tooltip = percent_format(accuracy = 1)(pct))) +
  facet_wrap(~tenure) +
  geom_col(position = position_fill(reverse = TRUE)) + 
  theme_pha() +
  scale_fill_manual(values = c("#f39152", "#a6cccc", "#e2e4e3")) +
  geom_col_interactive(position = position_fill(reverse = TRUE), size = 2) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = paste0(name, ": Share of cost-burdened households by tenure"),
       subtitle = "2015 to 2018",
       fill = "Cost burden",
       caption = "**Source:** U.S Department of Housing and Urban Development,\nComprehensive Housing Affordability Strategy (CHAS), Table 7.") +
  theme(legend.position = "top")


if (knitr::is_html_output()) {
  
  girafe(ggobj = cb_plot,
         height_svg = 4) 
  
} else {
  
  cb_plot
  
  }

```

Available data from William and Mary's Project HOPE shows that the number of enrolled students experiencing homelessness in the county has increased in recent years. The county's McKinney-Vento Act data showed that the number of students experiencing homeless has doubled from the 2016-2017 school year when the number was 15. By the 2019-2020, that number was 29.

```{r}
#| label: fig-mkv
#| fig-cap: !expr "paste0(name, \"Enrolled students experiencing homelessness by school year\")"

div <- 72

mkv <- read_csv("data/mkv.csv") |> 
  subset(Division %in% div)

mkv_plot <- ggplot(mkv,
       aes(x = year,
           y = students,
           fill = students,
           data_id = students, 
           tooltip = students)) +
  geom_col() + 
  geom_col_interactive(size = 2) +
  theme_pha() + 
  facet_wrap(~Name) + 
  scale_fill_gradientn(colours = c("#CAE3C2", "#74BA91", "#519B8D", "#2B4258")) +
  labs(title = "Enrolled students experiencing homelessness by school year",
       subtitle = "2016-2017 to 2019-2020",
       caption = "**Source:** William & Mary School of Education, Project HOME - Virginia.")

if (knitr::is_html_output()) {
  
  girafe(ggobj = mkv_plot,
         height_svg = 4) 
  
} else {
  
  mkv_plot
  
  }

```
