# (PART) Local summaries {.unnumbered}

# Richmond City {#part-4-1}

This chapter is a summary of the major changes to the City of Richmond's population and housing market in the past five years.

```{r setup}

library(tidyverse)
library(sf)
library(sp)
library(plotly)
library(simplevis)
library(tidytext)
library(scales)
library(readxl)
library(zoo)
library(lubridate)

years <- 2016:2020
rva <- 51760

```

## Takeaways

```
Add short bullet points summarizing major changes
```

## Demographic and socioeconomic changes

### Population changes

```{r rva-pop}

#Add choice of population stat (net change, components, etc)

rva_pop <- read_rds("data/pha_pop.rds") |> 
  filter(NAMELSAD == "Richmond city") |> 
  st_drop_geometry()

rva_pop_bar <- ggplot(rva_pop,
                      aes(x = year,
                          y = value)) +
  geom_col()

ggplotly(rva_pop_bar) |> 
  plotly_camera()
  

```

```{r rva-comp}

rva_change <- read_rds("data/pep_change_clean.rds") |> 
  filter(GEOID == rva)

pep_2021 <- read_rds("data/pep_2021.rds") |> 
  filter(GEOID == rva)

rva_change_21 <- rbind(rva_change, pep_2021)


```
### Household characteristics

```{r rva-children}
# Add choice of household stat (HHs with children, HHs with seniors, subfamilies, etc)

b25115_vars <- load_variables(2020, "acs5") |>
  filter(str_sub(name, end = 6) %in% "B25115")

b25115_raw <- map_dfr(years, function(yr){
  b25115_pull <- get_acs(
    geography = "county",
    state = "VA",
    table = "B25115",
    year = yr,
    survey = "acs5"
  ) |>
    mutate(year = yr)
})

b25115_raw <- b25115_raw |>
  subset(GEOID %in% rva)

b25115_vars_cleaned <- b25115_vars |>
  separate(label, into = c("est", "total", "tenure", "family", "type", "rel", "children"), sep = "!!") |>
  select(variable = name, tenure, family, type, rel, children) |>
  mutate(tenure = case_when(
    tenure == "Owner occupied:" ~ "Homeowner",
    tenure == "Renter occupied:" ~ "Renter"
  )) |>
  mutate(family = case_when(
    family == "Family households:" ~ "Family household",
    family == "Nonfamily households" ~ "Nonfamily household"
  )) |>
  mutate(type = case_when(
    type == "Married-couple family:" ~ "Married household",
    rel == "Male householder, no spouse present:" ~ "Single male household",
    rel == "Female householder, no spouse present:" ~ "Single female household",
    family == "Nonfamily household" ~ "Nonfamily household"
  )) |>
  mutate(children = case_when(
    rel == "With own children of the householder under 18 years" ~ "With children",
    rel == "No own children of the householder under 18 years" ~ "No children",
    children == "No own children of the householder under 18 years" ~ "No children",
    children == "With own children of the householder under 18 years" ~ "With children",
    family == "Nonfamily household" ~ "Nonfamily household"
  )) |>
  select(variable, tenure, family, type, children) |>
  drop_na()

rva_b25115 <- b25115_raw |> 
  right_join(b25115_vars_cleaned, by = "variable") |>
  select(NAME, GEOID, year, tenure, type, children, estimate) |> 
  pivot_wider(names_from = year,
              values_from = estimate) |>
  transform(change = `2020` - `2016`) |>
  select(tenure, type, children, change) |>
  group_by(tenure, children) |>
  summarise(change = sum(change))


ggplot(rva_b25115,
       aes(x = children, y = change, fill = tenure)) +
  geom_col(position = "dodge") + 
  facet_wrap(~children, nrow = 1, scales = "free_x") +
  scale_y_continuous(labels = label_comma(),
                     breaks = c(-3000,0,3000,6000,9000),
                     limits = c(-3050,9050)) +
  labs(title = "Change in households with children by tenure",
       subtitle = "2016 to 2020",
       caption = "Source: U.S. Census Bureau, American Community Survey, Table B25115.") +
  theme(axis.title = element_blank(),
        axis.text.x = element_blank(),
        panel.background = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(color = "grey95",
                                          size = 0.05),
        axis.ticks = element_blank()) +
  scale_fill_manual(values = tenure_colors)

```

```{r rva-seniors}

b09020_vars <- load_variables(2020, "acs5") |>
  filter(str_sub(name, end = 6) %in% "B09020")

b09020_raw <- map_dfr(years, function(yr){
  b09020_pull <- get_acs(
    geography = "county",
    state = "VA",
    table = "B09020",
    year = yr,
    survey = "acs5"
  ) |>
    mutate(year = yr)
})

b09020_raw <- b09020_raw |>
  subset(GEOID %in% rva)

b09020_vars_cleaned <- b09020_vars |>
  separate(label, into = c("est", "total", "hhgq", "family", "relationship", "sex", "alone"), sep = "!!") |>
  select(variable = name, hhgq, family, relationship, alone) |>
  filter(!variable %in% c("B09020_001", "B09020_002", "B09020_003", "B09020_005", "B09020_006",
                          "B09020_012", "B09020_013", "B09020_014", "B09020_017")) |>
  mutate(relationship = case_when(
    relationship == "Spouse" ~ "With spouse",
    relationship == "Nonrelatives" ~ "With nonrelatives",
    relationship %in% c("Parent", "Parent-in-law", "Other relatives") ~ "With other relative(s)",
    hhgq == "In group quarters" ~ "Group quarters",
    !is.na(alone) ~ alone,
    TRUE ~ relationship
  )) |>
  mutate(family = case_when(
    family == "In family households:" ~ "Family",
    family == "In nonfamily households:" ~ "Nonfamily",
    hhgq == "In group quarters" ~ "Group quarters"
  )) |>
  mutate(across(.fns = ~str_remove_all(.x, ":"))) |>
  select(1,3,4)

rva_b09020 <- b09020_raw |>
  right_join(b09020_vars_cleaned, by = "variable") |>
  select(NAME, GEOID, year, family, relationship, estimate) |>
  mutate(NAME = str_remove_all(NAME, ", Virginia")) |>
  group_by(year, family, relationship) |>
  summarise(estimate = sum(estimate)) |>
  group_by(family, relationship) |>
  mutate(diff = estimate - lag(estimate)) |>
  drop_na() |>
  mutate(run_diff = cumsum(diff))


rva_b09020 <- rva_b09020 |> 
  ungroup() |> 
  filter(year == 2020) |> 
  mutate(family = str_replace(family, "Group quarters", " "),
         family = fct_relevel(family, "Family", "Nonfamily", " "))

ggplot(rva_b09020,
       aes(y = reorder_within(relationship, run_diff, family),
           x = run_diff,
           fill = relationship)) +
  geom_col(position = "dodge") +
  scale_y_reordered() +
  scale_x_continuous(labels = label_comma(),
                     breaks = c(0,2000,4000,6000,8000)) +
  facet_grid(rows = vars(family), scales = "free_y", space = "free", switch = "y") +
  #scale_y_discrete(labels = function(x) str_wrap(x, width = 10)) +
  labs(title="Change in senior population by living arrangement",
       subtitle="2016 to 2020",
       fill="Tenure",
       caption = "Source: U.S. Census Bureau, American Community Survey, Table B09020.") +
  theme(axis.title = element_blank(),
        #axis.text.x = element_blank(),
        panel.background = element_blank(),
        panel.grid.major.y = element_blank(),
        legend.position = "none",
        panel.grid.major.x = element_line(color = "grey95",
                                          size = 0.05),
        axis.ticks = element_blank(),
        strip.placement = "outside",
        strip.text.y = element_text(angle = 0))

```
### Income and wages

```{r rva-income}

#Add choice of income stat (probably median household income of some kind)

years <- 2016:2020

b25119_vars <- load_variables(2020, "acs5") |> 
  filter(str_sub(name, end = 6) %in% "B25119")

b25119_raw <- map_dfr(years, function(yr){
  b25119_pull <- get_acs(
    geography = "county",
    state = "VA",
    table = "B25119",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) %>%
    mutate(year = yr)
  })

b25119_raw <- b25119_raw |> 
  subset(GEOID %in% rva)

b25119_vars_cleaned <- b25119_vars |> 
  separate(label, into = c("est", "income", "total", "tenure"), sep = "!!") |>  
  select(variable = name, tenure) |> 
  mutate(tenure = case_when(
    tenure == "Owner occupied (dollars)" ~ "Homeowner",
    tenure == "Renter occupied (dollars)" ~ "Renter"
  ))

b25119_vars_cleaned$tenure <- b25119_vars_cleaned$tenure |>  replace_na('All households')

rva_b25119 <- b25119_raw |> 
  right_join(b25119_vars_cleaned, by = "variable") |> 
  select(NAME, GEOID, year, tenure, estimate, moe)

cpi <- read_excel("data/CPI_U_RS.xlsx")
cpi <- cpi |> 
  rename(year = Year,
         priceindex = Index) |> 
  transform(year = as.numeric(year))

b25119_rva <- rva_b25119 |> 
    left_join(cpi, by = 'year') |> 
  transform(dollars20 = ((381.2/priceindex)*estimate)) |> 
  select(NAME, GEOID, year, tenure, dollars20, cdollars = estimate) |> 
  separate(NAME, into = c("locality", "state"), sep = ",") |> 
  filter(tenure != "All households") |> 
  select(year, locality, tenure, dollars20)

ggplot(b25119_rva, 
       aes(x = year,
           y = dollars20,
           color = tenure)) +
  geom_line(stat = "identity", size = 1) +
  theme(axis.title = element_blank()) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Median houshold income by tenure")

```

### Persons with disabilities


```{r ind-living}

# Add disability stat (probably independent living difficulty)

# Table B18107 Sex by age by independent living difficulty

# "Because of a physical, mental, or emotional problem, having difficulty doing errands alone such as visiting a doctor’s office or shopping (DOUT)."


b18107_vars <- load_variables(2020, "acs5") |> 
  filter(str_sub(name, end = 6) %in% "B18107")

b18107_raw <- map_dfr(years, function(yr){
  b10107_pull <- get_acs(
    geography = "county",
    state = "VA",
    table = "B18107",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) |> 
    mutate(year = yr)
  })

b18107_raw <- b18107_raw |> 
  subset(GEOID %in% rva)

b18107_vars_cleaned <- b18107_vars |> 
  separate(label, into = c("est", "total", "sex", "age", "indliv"), sep = "!!") |> 
  select(variable = name, sex , age, indliv) |> 
  drop_na() |> 
  mutate(across(.fns = ~str_remove_all(.x, ":")))

rva_b18107 <- b18107_raw |> 
  right_join(b18107_vars_cleaned, by = "variable") |> 
  select(NAME, GEOID, year, sex, age, indliv, estimate) |> 
  group_by(year, age, indliv) |> 
  summarise(estimate = sum(estimate)) |> 
  filter(indliv == "With an independent living difficulty") |> 
  pivot_wider(names_from = year,
              values_from =  estimate) |> 
  transform(change = `2020` - `2016`) |> 
  select(age, indliv, change)

ggplot(rva_b18107,
       aes(x = age,
           y = change)) +
  geom_bar(stat = "identity", fill = "blue") +
  coord_flip() +
  scale_x_discrete(labels = rev) +
  scale_y_continuous(labels = number_format(big.mark = ",")) +
  theme(axis.title = element_blank()) +
  labs(title = "Change in population with independent living difficulty by age: \n2016 to 2020")
  



```

## Housing supply and market changes

### Homeownership

```{r rva-sales}

# Median sales price over time

richmond_price <- read_csv("data/richmond_med_sales.csv") |> 
  clean_names() |> 
  mutate(geography = "Richmond City") |> 
  pivot_longer(cols = starts_with("sale"),
    names_to = "label",
    values_to = "med_price",
    values_drop_na = TRUE) |> 
  mutate(year = substr(label, 19,23)) |> 
  select(geography, year, month, med_price)

richmond_price$month <- match(richmond_price$month, month.abb)

richmond_price <- richmond_price |> 
  mutate(date = make_date(year, month)) |> 
  mutate(med_price = parse_number(med_price))

ggplot(richmond_price,
       aes(x = date,
           y = med_price)) +
  geom_line(stat = "identity") +
  scale_y_continuous(labels = dollar_format())

```

### Rental


```{r rva-rent-submarket}

# Average asking rent over time (by submarket if applicable, maybe by bedroom if no submarkets)

# Disaggregate by Richmond CoStar Submarket

# CoStar Filters: Construction Status - Existing

# Added locality field manually

rva_sub_rent <- read_csv("data/rr_submarket_data.csv") |> 
  clean_names() |> 
  separate(geography, into = c("region", "state", "submarket"), sep = " - ") |> 
  select(period, locality, submarket, "rent" = market_asking_rent_unit, vacancy_rate) |> 
  mutate(rent = parse_number(rent)) |> 
  mutate(period = as.Date(as.yearqtr(period, format = "%Y Q%q"), frac = 1)) |> 
  filter(locality == "Richmond City")

# Issue with above is that rents are not inflation-adjusted. May need to revisit. I've requested from CoStar to allow for an inflation-adjusted average rent per unit pull, BUT who knows if they can add it. Might consider just having a table with most recent rents.

rsub_rent_line <- ggplot(rva_sub_rent,
       aes(x = period,
           y = rent, 
           color = submarket)) +
  geom_line(stat = "identity") + 
  labs(title = "Average asking rent by submarket")
  

ggplotly(rsub_rent_line)

```


### Housing assistance

```
Change in number of assisted units (NHPD)
```

### Naturally-occuring affordable housing

```{r rva-noah}
# Add choice of NOAH stat (total number, change in rent, etc) 

# Probably do a histogram by age

rva_noah <- read_csv("data/pha_noah_81722.csv") |> 
  clean_names() |> 
  st_as_sf(coords = c("longitude", "latitude"), crs=4326, remove=FALSE) |> 
  filter(property_name != "Sedgefield" | property_name != "Holidy Mobile Home Park") |> 
  filter(county_name == "Richmond City")


```

## Gap analysis


```{r rva-rent-gap}

# Compare local median renter income to average asking rent

rva_income <- b25119_cpi |> 
  filter(locality == "Richmond city") |> 
  filter(tenure == "Renter")

rva_rent <- read_csv("data/rr_annual_rent.csv") |> 
    pivot_longer(cols = 2:8,
               names_to = "locality",
               values_to = "rent_current") |> 
  clean_names() |> 
  filter(locality == "Richmond City")

rva_rent_gap <- rva_income |> 
  left_join(rva_rent, by = "year") |> 
  mutate(rent_current = parse_number(rent_current)) |> 
  mutate(affordable_rent = (cdollars/12)*.30) |> 
  mutate(difference = rent_current - affordable_rent) |> 
  select(year, rent_current, affordable_rent) |> 
  pivot_longer(2:3,
               names_to = "type",
               values_to = "dollars")

rrent_gp <- ggplot(rva_rent_gap,
       aes(x = year,
           y = dollars,
           color = type
        )) +
  geom_line(stat = "identity", size = 1) +
  scale_y_continuous(labels = dollar_format()) +
  theme(axis.title = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1)) +
  labs(title = "Average asking rent versus rent affordable to typical renter", color = "Nominal rent")
  
ggplotly(rrent_gp) |> 
  plotly_camera()

```


```{r  rva-own-gap}

#Compare renter income versus home prices

rr_sales_annual <- read_csv("data/rr_annual_sales.csv") |> 
  pivot_longer(cols = 2:9,
               names_to = "locality",
               values_to = "price") |> 
  clean_names() |> 
  filter(locality == "Richmond City")

int_annual <- read_csv("data/fredmac_int_annual.csv")

downpayment <- 0.05 # 5% downpayment
closingcosts <- 0.015 # 1.5% closing costs
utilities <- 250 # Assume $250/month for utilities

rva_sale <- rr_sales_annual |> 
  left_join(rva_income, by = "year") |> 
  left_join(int_annual, by = "year") |> 
  drop_na() |> 
  mutate(principal = price - (price*downpayment)) |> 
  mutate(loanamt = principal/(1-closingcosts)) |> 
  mutate(mortgage = abs(PMT((annual_int/12), 360, loanamt)) + 250) |> 
  mutate(inc_needed = ((mortgage*10)/2.8)*12) |> 
  select(year, inc_needed, cdollars)

rva_slgap <- rva_sale |> 
  pivot_longer(2:3,
               names_to = "data",
               values_to = "value") |> 
  mutate(data = case_when(
    data == "cdollars" ~ "Renter median household income",
    data == "inc_needed" ~ "Income needed to afford median home price"
  ))

rsale_gp <- ggplot(rva_slgap,
       aes(x = year, 
           y = value,
           color = data)) +
  geom_line(stat = "identity", size = 1) +
  labs(title = "Income needed to afford median home price versus median renter income", color = "Nominal dollars") +
  theme(axis.title = element_blank()) +
  scale_y_continuous(labels = dollar_format())

ggplotly(rsale_gp) |> 
  plotly_camera()

```
...

### Affordability of current housing stock

```{r rva-gap}
# Add choice of affordability stat (surplus/deficit, income vs prices, etc)

rva_gap <- read_rds("data/tb_18_match.rds") |> 
  filter(county == "Richmond city") |> 
  group_by(year, household_income, gapcode) |>
  summarise(estimate = sum(estimate)) |>
  mutate(estimate = case_when(
    gapcode == "Gap" ~ estimate * -1,
    gapcode == "Matches or less than income" ~ estimate
  )) |>
  filter(household_income != "80 percent AMI or greater")

rva_gap_bar <- ggplot(rva_gap,
       aes(x = year, 
           y = estimate,
           fill = gapcode)) +
  geom_bar(stat = "identity") +
  facet_grid(~household_income) +
  theme(axis.title = element_blank()) +
  labs(title = "Rental housing gap by AMI: 2015 to 2018", fill = "Gap/Affordability")

ggplotly(rva_gap_bar) |> 
  plotly_camera()

```

### Impact of housing costs


```{r rva-cb}

cb_7 <- read_rds("data/cb_7.rds")

cb_rva <- cb_7 |> 
  filter(county == "Richmond city",
         cb_group == "Cost-burdened") |> 
  group_by(year, county, tenure) |> 
  summarise(estimate = sum(estimate)) |> 
  group_by(county, tenure) |> 
  mutate(change = estimate - lag(estimate)) |> 
  drop_na() |> 
  mutate(run_diff = cumsum(change))
  
ggplot(cb_rva,
       aes(x = year, y = run_diff, fill = tenure)) +
  geom_col(position = "dodge") +
  facet_wrap(vars(county)) +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Cumulative change in cost-burdened households by tenure",
       subtitle = "2015 to 2018",
       fill = "Tenure",
       caption = "Source: U.S Department of Housing and Urban Development,\nComprehensive Housing Affordability Strategy (CHAS), Table 7.") +
  theme(axis.title = element_blank())


```
Add choice of housing cost stat (cost burden, PIT, eviction filings, etc)
```{r rva-evictions}

rva_evictions <- read_csv("data/evictions_17_22.csv") |> 
  pivot_longer(3:128,
               names_to = "data",
               values_to = "value") |> 
  mutate(type = case_when(
    str_detect(data, "Filings") ~ "Filings",
    str_detect(data,"Evictions") ~ "Eviction judgement"
  )) |> 
  mutate(date = str_sub(data, -7, -1)) |> 
  mutate(date = ym(date)) |> 
  mutate(year = year(date)) |> 
  mutate(type = fct_relevel(type, "Filings", "Eviction orders")) |> 
  filter(Jurisdiction == "Richmond City")

eviction_colors <- setNames(c("#00BFC4", "#F8766D"), nm = c("Filings", "Eviction orders"))

ggplot(rva_evictions,
       aes(x = date,
           y = value,
           fill = type)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Evictions filings and orders by locality",
       subtitle = "January 2017 to March 2022",
       caption = "Source: RVA Eviction Lab") +
  scale_fill_manual(values = eviction_colors) +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        axis.title = element_blank(),
        panel.background = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(color = "grey95",
                                          size = 0.05),
        axis.ticks = element_blank())

```
