# Town of Ashland {#part-4-9}

This chapter is a summary of the major changes to the Town of Ashland's population and housing market in the past five years.

```{r}
#| label: setup

library(tidyverse)
library(tidycensus)
library(sf)
library(sp)
library(plotly)
library(simplevis)
library(tidytext)
library(scales)
library(readxl)
library(zoo)
library(lubridate)
library(janitor)
library(viridis)
library(kableExtra)
library(leaflet)
library(optiRum)

tenure <- c("Homeowner", "Renter")
tenure_colors <-setNames(viridis(2), nm = tenure)

```

## Takeaways


## Demographic and socioeconomic changes

### Population changes

The gradual population growth that Ashland was seeing throughout the last decade took a change in course between 2019 and 2020. Between these years, it was estimated that Ashland's population declined by 310 individuals. This was a greater decrease than the town's last recorded population decrease between 2010 and 2011, when the town population declined by 38. The small population of Ashland may warrant some caution in describing trends between American Community Survey estimates and decennial census counts.

```{r}
#| label: pop-data
#| echo: false


ashland_pop <- get_estimates(
  geography = "place",
  state = "VA",
  variables = "POP",
  year = 2019,
  time_series = TRUE
) %>% 
  filter(NAME == "Ashland town, Virginia")

ashland_20 <- get_decennial(
  geography = "place",
  state = "VA",
  year = 2020,
  sumfile = "pl",
  variables = "P1_001N"
) %>% 
  filter(NAME == "Ashland town, Virginia")

ashpop_clean <- ashland_pop |>
  filter(!DATE %in% c(2, 3)) |> # Remove non-decennial 2010 counts
  mutate(year = # Translate date codes into years
    case_when(
      DATE == 1 ~ "2010",
      DATE == 4 ~ "2011",
      DATE == 5 ~ "2012",
      DATE == 6 ~ "2013",
      DATE == 7 ~ "2014",
      DATE == 8 ~ "2015",
      DATE == 9 ~ "2016",
      DATE == 10 ~ "2017",
      DATE == 11 ~ "2018",
      DATE == 12 ~ "2019")) |> 
  mutate(counttype = # Add descriptions to count types
      case_when(
        DATE == 1 ~ "Census population",
        TRUE ~ "Population estimate")) |> 
  select( # Simplify columns
    GEOID,
    year,
    counttype,
    value
  )

# Prep total population counts from 2020 Census summary file

census_clean <- ashland_20 |> 
  mutate(year = "2020", # Add year and count type columns
         counttype = "Census population") |>
  select( # Simplify columns
    GEOID,
    year,
    counttype,
    value
  )

population_data <- ashpop_clean |> 
  bind_rows(census_clean)

```

```{r}
#| label: fig-ash-pop
#| fig-cap: "Town of Ashland: Total Population"

 ggplot(population_data,
        aes(x = year,
            y = value,
            fill = value)) +
  geom_col() +
  scale_fill_continuous(viridis_pal()) +
  labs(title = "Town of Ashland: Total Population",
       subtitle = "2010 to 2020",
       caption = "Source: U.S. Census Bureau Decennial Census and American Community Survey.") +
  scale_y_continuous(labels = label_comma()) +
  theme(axis.title = element_blank(),
        legend.position = "none",
        panel.background = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major.y = element_line(color = "grey95",
                                          size = 0.05))


```
Population projections produced by the University of Virginia's Weldon Cooper Center show that the town will continue to grow over the coming decades. By 2050, the Weldon Cooper Center expects Ashland to approach nearly 10,000 residents --- a 27 percent increase over 30 years.

```{r}
#| label: pop-forecasts

years <- c(2020, 2030, 2040, 2050)
estimate <- c(7565, 8142, 8824, 9638)
type <- c("Census", "Forecast", "Forecast", "Forecast")

forecast <- data.frame(year = years,
                       estimate = estimate,
                       type = type)


ggplot(forecast,
       aes(x = year,
           y = estimate,
           fill = type,
           label = label_comma()(estimate))) +
  geom_col() +
  geom_text(vjust= 1.5,
            color="white",
            size = 3,
            fontface="bold") +
  scale_y_continuous(labels = label_comma()) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major.y = element_line(color = "grey95",
                                          size = 0.05),
        panel.background = element_blank()) +
  labs(title="Town of Ashland: Population projections",
       subtitle="Town Population in 2020 and forecast to 2050",
       fill="Estimate type",
       caption="Source: University of Virginia Weldon Cooper Center for Public Service")


  

```


### Household characteristics

From 2016 to 2020, Ashland has seen most of its household growth among homeowners between the ages of 25 and 44 years old (+342). Declines have mainly been among homeowner householders older than 45 years old (-249), while there has been an increase of renter households among this same age group.

```{r}
#| label: hh-age-data
#| eval: FALSE

years <- 2016:2020

b25007_vars <- load_variables(2020, "acs5") |>
  filter(str_sub(name, end = 6) %in% "B25007")

b25007_raw <- map_dfr(years, function(yr){
  b25007_pull <- get_acs(
    geography = "place",
    state = "VA",
    table = "B25007",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) |>
    mutate(year = yr)
}) %>% 
  filter(NAME == "Ashland town, Virginia")


b25007_vars_cleaned <- b25007_vars |>
  separate(label, into = c("est", "total", "tenure", "age"), sep = "!!") |>
  select(variable = name, tenure, age) |>
  drop_na() |>
  mutate(tenure = case_when(
    tenure == "Owner occupied:" ~ "Homeowner",
    tenure == "Renter occupied:" ~ "Renter"
  ))

b25007_data <- b25007_raw |>
  right_join(b25007_vars_cleaned, by = "variable") |>
  select(NAME, GEOID, year, tenure, age, estimate, moe) |>
  mutate(NAME = str_remove_all(NAME, "town, Virginia")) |>
  mutate(age = case_when(
    age == "Householder 15 to 24 years" ~ "Under 25 years old",
    age == "Householder 25 to 34 years" ~ "25 to 44 years old",
    age == "Householder 35 to 44 years" ~ "25 to 44 years old",
    age == "Householder 45 to 54 years" ~ "45 to 64 years old",
    age == "Householder 55 to 59 years" ~ "45 to 64 years old",
    age == "Householder 60 to 64 years" ~ "45 to 64 years old",
    age == "Householder 65 to 74 years" ~ "65 years and over",
    age == "Householder 75 to 84 years" ~ "65 years and over",
    age == "Householder 85 years and over" ~ "65 years and over"
  )) |>
  group_by(year, tenure, age) |>
  summarise(estimate = sum(estimate)) |>
  pivot_wider(names_from = year,
              values_from = estimate) |>
  select(tenure, age, `2016`, `2020`) |>
  transform(change = `2020` - `2016`) |>
  group_by(tenure, age) |>
  summarise(change = sum(change)) |>
  mutate(age = fct_relevel(age, "Under 25 years old")) |> 
  group_by(tenure, age) |> 
  summarise(change = sum(change))
```

```{r}
#| label: fig-hh-age-plot
#| fig.cap: "Change in households by age and tenure"

ggplot(b25007_data,
       aes(y = change, x =tenure, fill = tenure)) +
  geom_col(position = "dodge") +
  facet_wrap(~age, nrow = 1) +
  scale_y_continuous(labels = label_comma()) +
  labs(title="Change in households by age and tenure",
       subtitle="2016 to 2020",
       fill="Tenure",
       caption = "Source: U.S. Census Bureau, American Community Survey, Table B25007.") +
  theme(axis.title = element_blank(),
        axis.text.x = element_blank(),
        panel.background = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(color = "grey95",
                                          size = 0.05),
        axis.ticks = element_blank()) +
  scale_fill_manual(values = tenure_colors)

```
Households with children have only increased among renters in Hanover (+83) from 2016 to 2020. But the greatest household type to see growth in the town has been among nonfamily homeowner households (i.e. individuals living alone or with )
```{r}
#| label: hh-child-data
#| eval: FALSE
#| 
years <- 2016:2020

b25115_vars <- load_variables(2020, "acs5") |>
  filter(str_sub(name, end = 6) %in% "B25115")

b25115_raw <- map_dfr(years, function(yr){
  b25115_pull <- get_acs(
    geography = "place",
    state = "VA",
    table = "B25115",
    year = yr,
    survey = "acs5"
  ) |>
    mutate(year = yr)
}) %>% 
  filter(NAME == "Ashland town, Virginia")


b25115_vars_cleaned <- b25115_vars |>
  separate(label, into = c("est", "total", "tenure", "family", "type", "rel", "children"), sep = "!!") |>
  select(variable = name, tenure, family, type, rel, children) |>
  mutate(tenure = case_when(
    tenure == "Owner occupied:" ~ "Homeowner",
    tenure == "Renter occupied:" ~ "Renter"
  )) |>
  mutate(family = case_when(
    family == "Family households:" ~ "Family household",
    family == "Nonfamily households" ~ "Nonfamily household"
  )) |>
  mutate(type = case_when(
    type == "Married-couple family:" ~ "Married household",
    rel == "Male householder, no spouse present:" ~ "Single male household",
    rel == "Female householder, no spouse present:" ~ "Single female household",
    family == "Nonfamily household" ~ "Nonfamily household"
  )) |>
  mutate(children = case_when(
    rel == "With own children of the householder under 18 years" ~ "With children",
    rel == "No own children of the householder under 18 years" ~ "No children",
    children == "No own children of the householder under 18 years" ~ "No children",
    children == "With own children of the householder under 18 years" ~ "With children",
    family == "Nonfamily household" ~ "Nonfamily household"
  )) |>
  select(variable, tenure, family, type, children) |>
  drop_na()

b25115_data <- b25115_raw |>
  right_join(b25115_vars_cleaned, by = "variable") |>
  select(GEOID, year, tenure, type, children, estimate) |>
  pivot_wider(names_from = year,
              values_from = estimate) |>
  transform(change = `2020` - `2016`) |>
  select(tenure, type, children, change) |>
  group_by(tenure, children) |>
  summarise(change = sum(change)) 

```

```{r}
#| label: fig-hh-child-plot
#| fig.cap: "Change in households with children by tenure"

ggplot(b25115_data,
       aes(x = children, y = change, fill = tenure)) +
  geom_col(position = "dodge") + 
  facet_wrap(~children, nrow = 1, scales = "free_x") +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Change in households with children by tenure",
       subtitle = "2016 to 2020",
       fill = "Tenure",
       caption = "Source: U.S. Census Bureau, American Community Survey, Table B25115.") +
  theme(axis.title = element_blank(),
        axis.text.x = element_blank(),
        panel.background = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(color = "grey95",
                                          size = 0.05),
        axis.ticks = element_blank()) +
  scale_fill_manual(values = tenure_colors)

```

```{r}
#| label: hh-seniors-data
#| eval: FALSE

years <- 2016:2020

b09020_vars <- load_variables(2020, "acs5") |>
  filter(str_sub(name, end = 6) %in% "B09020")

b09020_raw <- map_dfr(years, function(yr){
  b09020_pull <- get_acs(
    geography = "place",
    state = "VA",
    table = "B09020",
    year = yr,
    survey = "acs5"
  ) |>
    mutate(year = yr)
}) %>% 
  filter(NAME == "Ashland town, Virginia")

b09020_vars_cleaned <- b09020_vars |>
  separate(label, into = c("est", "total", "hhgq", "family", "relationship", "sex", "alone"), sep = "!!") |>
  select(variable = name, hhgq, family, relationship, alone) |>
  filter(!variable %in% c("B09020_001", "B09020_002", "B09020_003", "B09020_005", "B09020_006",
                          "B09020_012", "B09020_013", "B09020_014", "B09020_017")) |>
  mutate(relationship = case_when(
    relationship == "Spouse" ~ "With spouse",
    relationship == "Nonrelatives" ~ "With nonrelatives",
    relationship %in% c("Parent", "Parent-in-law", "Other relatives") ~ "With other relative(s)",
    hhgq == "In group quarters" ~ "Group quarters",
    !is.na(alone) ~ alone,
    TRUE ~ relationship
  )) |>
  mutate(family = case_when(
    family == "In family households:" ~ "Family",
    family == "In nonfamily households:" ~ "Nonfamily",
    hhgq == "In group quarters" ~ "Group quarters"
  )) |>
  mutate(across(.fns = ~str_remove_all(.x, ":"))) |>
  select(1,3,4)

b09020_data <- b09020_raw |>
  right_join(b09020_vars_cleaned, by = "variable") |>
  select(GEOID, year, family, relationship, estimate) |>
  group_by(year, family, relationship) |> 
  summarise(estimate = sum(estimate)) |> 
  pivot_wider(names_from = year,
              values_from = estimate) |>
  transform(change = `2020` - `2016`) |>
  select(family, relationship, change) |>
  group_by(family, relationship) |> 
  summarise(change = sum(change)) |> 
  ungroup() |>
  mutate(family = str_replace(family, "Group quarters", " "),
         family = fct_relevel(family, "Family", "Nonfamily", " "))

```
While the senior population has grown overall in Ashland (+107), that growth has been largely among seniors who live with other relatives (+225) or with a spouse (+52). Seniors who are head of households has declined in the town (-87), as well as the number of seniors living in group quarters (-69).

```{r}
#| label: fig-hh-seniors-plot
#| fig.cap: "Change in senior population by living arrangement"

ggplot(b09020_data,
       aes(y = reorder_within(relationship, change, family),
           x = change,
           fill = relationship)) +
  geom_col(position = "dodge") +
  scale_y_reordered() +
  scale_x_continuous(labels = label_comma()) +
  facet_grid(rows = vars(family), scales = "free_y", space = "free", switch = "y") +
  labs(title="Change in senior population by living arrangement",
       subtitle="2016 to 2020",
       fill="Tenure",
       caption = "Source: U.S. Census Bureau, American Community Survey, Table B09020.") +
  theme(axis.title = element_blank(),
        #axis.text.x = element_blank(),
        panel.background = element_blank(),
        panel.grid.major.y = element_blank(),
        legend.position = "none",
        panel.grid.major.x = element_line(color = "grey95",
                                          size = 0.05),
        axis.ticks = element_blank(),
        strip.placement = "outside",
        strip.text.y = element_text(angle = 0))

```

### Income and wages

Household incomes have also shifted in Ashland. There has been a larger increase among renter households making more than \$50,000 than homeowners in the same income group. The greatest increase among homeowners has occurred with homeowners making between \$15,000 and \$49,999. Households making under \$15,000 have decreased for both homeowners and renters. 

```{r}
#| label: med-inc
#| eval: false

years <- 2016:2020

b25119_vars <- load_variables(2020, "acs5") %>% 
 filter(str_sub(name, end = 6) %in% "B25119")

b25119_raw <- map_dfr(years, function(yr){
  b25119_pull <- get_acs(
    geography = "place",
    state = "VA",
    table = "B25119",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) |> 
    mutate(year = yr)
  }) %>% 
  filter(NAME == "Ashland town, Virginia")

b25119_vars_cleaned <- b25119_vars %>% 
  separate(label, into = c("est", "lab", "total", "tenure"), sep = "!!") %>% 
  select(variable = name, tenure) %>% 
  drop_na() %>% 
  mutate(tenure = case_when(
    tenure == "Owner occupied (dollars)" ~ "Homeowner",
    tenure == "Renter occupied (dollars)" ~ "Renter"
  ))

b25119_data <- b25119_raw %>% 
  right_join(b25119_vars_cleaned, by = "variable") %>% 
  select(year, tenure, estimate)

```

```{r}
#| label: hh-inc-data
#| eval: FALSE


years <- 2016:2020

b25118_vars <- load_variables(2020, "acs5") |> 
  filter(str_sub(name, end = 6) %in% "B25118")

b25118_raw <- map_dfr(years, function(yr){
  b25118_pull <- get_acs(
    geography = "place",
    state = "VA",
    table = "B25118",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) |> 
    mutate(year = yr)
  }) %>% 
  filter(NAME == "Ashland town, Virginia")

b25118_vars_cleaned <- b25118_vars |> 
  separate(label, into = c("est", "total", "tenure", "income"), sep = "!!") |>  
  select(variable = name, tenure, income) |> 
  mutate(tenure = case_when(
    tenure == "Owner occupied:" ~ "Homeowner",
    tenure == "Renter occupied:" ~ "Renter"
  )) |> 
  drop_na()

b25118_data <- b25118_raw |> 
  right_join(b25118_vars_cleaned, by = "variable") |> 
  select(year, tenure, income, estimate) |> 
  mutate(income = case_when(
    income == "Less than $5,000" ~ "Less than $15,000",
    income == "$5,000 to $9,999" ~ "Less than $15,000",
    income == "$10,000 to $14,999" ~ "Less than $15,000",
    income == "$15,000 to $19,999" ~ "$15,000 to $24,999",
    income == "$$20,000 to $24,999" ~ "$15,000 to $24,999",
    income == "$25,000 to $34,999" ~ "$25,000 to $49,999",
    income == "$35,000 to $49,999" ~ "$25,000 to $49,999",
    income == "$50,000 to $74,999" ~ "$50,000 to $74,999",
    TRUE ~ income
  )) |> 
  group_by(year, tenure, income) |> 
  summarise(estimate = sum(estimate)) |> 
  filter(year %in% c(2016, 2020)) |> 
  pivot_wider(names_from = year,
              values_from = estimate) |>
  mutate(change = `2020` - `2016`,
         pct_change = change/`2016`)

```

```{r}
#| label: fig-hh-inc-plot
#| fig.cap: "Change in households by tenure and income level"

ggplot(b25118_data, 
       aes(y = income,
           x = change,
           fill = tenure)) +
  geom_col(position = "dodge") +
  facet_grid(~tenure) +
  scale_x_continuous(labels = label_comma()) +
  scale_y_discrete(limits = c("Less than $15,000","$15,000 to $24,999",
                              "$25,000 to $49,999", "$50,000 to $74,999",
                              "$75,000 to $99,999", "$100,000 to $149,999",
                              "$150,000 or more"),
                   labels = function(x) str_wrap(x, width = 12)) +
  labs(title = "Change in households by tenure and income level",
       subtitle="2016 to 2020",
       caption = "Source: U.S. Census Bureau, American Community Survey, Table B25118.") +
  theme(axis.title = element_blank(),
        panel.background = element_blank(),
        panel.grid.major.y = element_blank(),
        legend.position = "none",
        panel.grid.major.x = element_line(color = "grey95",
                                          size = 0.05),
        axis.ticks = element_blank()) +
  scale_fill_manual(values = tenure_colors)

```

### Persons with disabilities

There has been a net increase of 32 individuals with independent living difficulties in Ashland. This increase has mostly been among individuals 18 to 34 years of age, while decreases were only experienced by individuals between 35 and 64 years old --- indicating a greater need for accessibility and assistance for young adults and seniors in the town.

```{r}
#| label: ind-liv-age-data
#| eval: FALSE

# Table B18107 Sex by age by independent living difficulty

# "Because of a physical, mental, or emotional problem, having difficulty doing errands alone such as visiting a doctor’s office or shopping (DOUT)."

years <- 2016:2020

b18107_vars <- load_variables(2020, "acs5") |> 
  filter(str_sub(name, end = 6) %in% "B18107")

b18107_raw <- map_dfr(years, function(yr){
  b10107_pull <- get_acs(
    geography = "place",
    state = "VA",
    table = "B18107",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) |> 
    mutate(year = yr)
  }) %>% 
  filter(NAME == "Ashland town, Virginia")

b18107_vars_cleaned <- b18107_vars |> 
  separate(label, into = c("est", "total", "sex", "age", "indliv"), sep = "!!") |> 
  select(variable = name, sex , age, indliv) |> 
  drop_na() |> 
  mutate(across(.fns = ~str_remove_all(.x, ":")))

b18107_data <- b18107_raw |> 
  right_join(b18107_vars_cleaned, by = "variable") |> 
  select(GEOID, year, sex, age, indliv, estimate) |> 
  group_by(year, age, indliv) |> 
  summarise(estimate = sum(estimate)) |> 
  filter(indliv == "With an independent living difficulty") |> 
  pivot_wider(names_from = year,
              values_from =  estimate) |> 
  transform(change = `2020` - `2016`) |> 
  group_by(age) |> 
  summarise(change = sum(change))

```



```{r}
#| label: fig-ind-liv-age-plot
#| fig.cap: "Net change in individuals with independent living difficulties by age"

ggplot(b18107_data,
       aes(x = age,
           y = change,
           fill = age)) +
  geom_col() +
  labs(title="Net change in individuals with independent living difficulties by age",
       subtitle="2016 to 2020",
       caption="Source: U.S. Census Bureau, American Community Survey, Table B18107.") +
  scale_y_continuous(labels = label_comma()) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none",
        panel.background = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(color = "grey95",
                                          size = 0.05))

```
## Housing supply and market changes

### Homeownership

```{r}

ashland <- st_read('https://services2.arcgis.com/sKZWgJlU6SekCzQV/arcgis/rest/services/Ashland_Boundary/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson', quiet = TRUE)

leaflet(ashland) %>% 
  addTiles() %>% 
  addPolygons()

# Data pulled on 2022-10-13

annual_cpi <- read_csv("data/CPIAUCSL_annual.csv") %>% 
  mutate(year = year(DATE)) %>% 
  select(year, cpi = CPIAUCSL)

ashland_medsales <- read_csv("data/ashland_med_sales.csv") %>% 
  clean_names() %>% 
  pivot_longer(cols = starts_with("sale"),
    names_to = "label",
    values_to = "med_price",
    values_drop_na = TRUE
  ) |> 
mutate(year = substr(label, 19,23)) %>% 
  select(year, month, med_price)

ashland_medsales$month <- match(ashland_medsales$month, month.abb)

ashland_medsales <- ashland_medsales %>% 
  mutate(date = make_date(year, month)) %>% 
  mutate(med_price = parse_number(med_price))

cpi <- read_csv("data/CPIAUCSL.csv") %>% 
  select(date = DATE, cpi = CPIAUCSL)

ashland_priceadj <- ashland_medsales %>% 
  left_join(cpi, by = "date") %>% 
  mutate(dollars = ((295.271/cpi)*med_price)) %>%  
  mutate(year = year(date)) %>% 
  select(date, year, med_price, dollars, cpi) %>% 
  filter(year >= 2017)

```

```{r}

ggplot(ashland_priceadj,
       aes(x = date, y = dollars, color = dollars)) +
  geom_line(size = 1) +
  scale_y_continuous(labels = label_dollar()) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(title = "Monthly median sales price",
       subtitle = "January 2017 to July 2022",
       caption = "Source: Central Virginia Regional Multiple Listing Service.") +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none",
        panel.background = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(color = "grey95",
                                          size = 0.05)) +
  scale_color_continuous(type = "viridis", label = dollar_format()) + 
  geom_smooth(method = "lm")

```

### Rental

```{r}

rent <- read_csv("data/ashland_rent.csv") %>%  
  pivot_longer(2:3,
               names_to = "rent",
               values_to = "value") %>%  
  mutate(period = as.Date(as.yearqtr(period, format = "%Y Q%q"), frac = 1)) %>%  
  mutate(year = year(period)) %>%  
  filter(year >= 2016) %>%  
  mutate(rent = case_when(
    rent == "rent" ~ "Nominal rent",
    rent == "rent_adj" ~ "Inflation-adjusted rent"
  ))

ggplot(rent,
       aes(x = period,
           y = value, 
           color = rent)) +
  geom_line(stat = "identity", size = 1) + 
  labs(title = "Town of Ashland: Average asking rent by submarket",
       subtitle = "2016 Q1 to 2022 Q3",
       color = "Dollar value") +
  scale_y_continuous(labels = label_dollar()) +
  scale_x_date(date_labels = "%Y", date_breaks = "2 years", limits = as.Date(c("2016-01-01", "2022-09-30"))) +
  theme(axis.title = element_blank(),
        panel.background = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(color = "grey95",
                                          size = 0.05))

```


```{r}

nhpd_coded_rr <- read_rds("data/nhpd_coded_rr.rds") |> 
  filter(locality == "Hanover County") %>% 
  filter(city == "ASHLAND")

```


```{r}
mhc <- st_read("https://services1.arcgis.com/Hp6G80Pky0om7QvQ/arcgis/rest/services/Mobile_Home_Parks/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson", quiet = TRUE) |> 
  filter(STATE == "VA",
         CITY == "ASHLAND")

library(leaflet)

leaflet(mhc) |> 
  addTiles() |> 
  addCircleMarkers()
```
```{r}

mhc_list <- read_csv("data/mhccv_data.csv") %>% 
  filter(`Park Name` == "Palm Leaf Mobile Home Park" | `Park Name` == "Sedgefield Manufactured Home")


```

```{r}

years <- 2016:2020

b25119_vars <- load_variables(2020, "acs5") %>% 
 filter(str_sub(name, end = 6) %in% "B25119")

b25119_raw <- map_dfr(years, function(yr){
  b25119_pull <- get_acs(
    geography = "place",
    state = "VA",
    table = "B25119",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) |> 
    mutate(year = yr)
  }) %>% 
  filter(NAME == "Ashland town, Virginia")

b25119_vars_cleaned <- b25119_vars %>% 
  separate(label, into = c("est", "lab", "total", "tenure"), sep = "!!") %>% 
  select(variable = name, tenure) %>% 
  drop_na() %>% 
  mutate(tenure = case_when(
    tenure == "Owner occupied (dollars)" ~ "Homeowner",
    tenure == "Renter occupied (dollars)" ~ "Renter"
  ))

b25119_data <- b25119_raw %>% 
  right_join(b25119_vars_cleaned, by = "variable") %>% 
  select(year, tenure, estimate)

ashland_rent <- read_csv("data/ashland_annual_rent.csv")

rentgap <- b25119_data %>% 
  filter(tenure == "Renter") %>% 
  left_join(ashland_rent, by = "year") %>% 
  mutate(affordable_rent = (estimate/12)*.30) %>% 
  select(year, rent, affordable_rent) %>% 
  pivot_longer(2:3, 
               names_to = "label",
               values_to = "rent") %>% 
  mutate(label = case_when(
    label == "rent" ~ "Current rent",
    label == "affordable_rent" ~ "Affordable rent"
  ))

ggplot(rentgap,
       aes(x = year,
           y = rent,
           fill = label)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "Town of Ashland: Average asking rent versus\nrent affordable to median renter income",
       subtitle = "2016 to 2020",
       fill = "Nominal rent",
       caption = "Sources: U.S. Census Bureau, American Community Survey, Table B25119\nand CoStar Group Inc.") +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank(),
        panel.grid.major.y = element_line(color = "grey95",
                                          size = 0.05))




```

```{r}


ashland_medsales <- read_csv("data/ashland_mls_medsales.csv")

int_annual <- read_csv("data/fredmac_int_annual.csv")

downpayment <- 0.05 # 5% downpayment
closingcosts <- 0.015 # 1.5% closing costs
utilities <- 250 # Assume $250/month for utilities

slgap <- b25119_data %>% 
  filter(tenure == "Renter") %>% 
  left_join(ashland_medsales, by = "year") %>% 
  left_join(int_annual, by = "year") |> 
  drop_na() |> 
  mutate(principal = price - (price*downpayment)) |> 
  mutate(loanamt = principal/(1-closingcosts)) |> 
  mutate(mortgage = abs(PMT((annual_int/12), 360, loanamt)) + 250) |> 
  mutate(inc_needed = ((mortgage*10)/2.8)*12) |> 
  select(year, inc_needed, estimate) %>% 
  pivot_longer(2:3,
               names_to = "label",
               values_to = "dollars") |> 
  mutate(label = case_when(
    label == "estimate" ~ "Renter median household income",
    label == "inc_needed" ~ "Income needed to afford median home price"
  ))
  
ggplot(slgap,
       aes(x = year,
           y = dollars,
           fill = str_wrap(label, 25))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "Town of Ashland: Income needed to afford median home price\nversus median renter income",
       subtitle = "2016 to 2020",
       fill = "Nominal income",
       caption = "Sources: U.S. Census Bureau, American Community Survey, Table B25119\nand Central Virginia Regional Multiple Listing Service.") +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank(),
        panel.grid.major.y = element_line(color = "grey95",
                                          size = 0.05),
        legend.key.height=unit(1, "cm"))
    

```

```{r}

ashland_cb <- read_csv("data/ashland_cb.csv")

