---
format:
  revealjs:
    incremental: true
    self-contained: true
    theme: simple
    height: 900
    width: 1400
    slide-number: true
title: "Richmond Regional Housing Framework Data Update"
subtitle: "Selected draft findings"
author: "HDAdvisors"
date: 2022-09-08
---

## Research scope

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(scales)
library(viridis)

pha_names <- c("Richmond city", "Hanover County", "Henrico County", "Chesterfield County")

pha <- c("51085", "51760", "51087", "51041")

geo_colors <- setNames(viridis(4), nm = pha_names)

```



*How have housing needs changed across the region since the original Framework release?*

. . .

Four parts:

1.  Demographic and socioeconomic changes
2.  Housing supply and market changes
3.  Gap analysis
4.  Local summaries

## Data source updates

| Source                                       | Original                   | Current                               |
|-----------------------------|------------------|-------------------------|
| American Community Survey                    | 2013-2017 5-year estimates | 2016-2020 5-year estimates            |
| Comprehensive Housing Affordability Strategy | 2012-2016 5-year estimates | 2014-2018 5-year estimates            |
| Census Population Estimates Program          | 2018                       | 2021, including 2020 Decennial Census |
| Census Building Permits Survey               | 2018                       | 2022, through July                    |
| MLS                                          | 2019                       | 2022, through August                  |
| CoStar                                       | *Not used*                 | 2022, through August                  |

## Other updated datasets

Additional sources used for this refresh include:

-   National Housing Preservation Database
-   BLS Occupational Employment and Wage Statistics
-   CFPB Mortgage Performance Trends
-   Point-in-Time homelessness counts and McKinney-Vento student homelessness statistics
-   District court eviction filings and judgments

# Demographic and socioeconomic changes

## More people come here than are born here.

<br>

```{r}
#| label: components-plot
#| echo: false

pha_comp_change <- read_rds("../data/rr_comp_change.rds") |> 
  filter(NAMELSAD %in% pha_names)

comp_chg_summary <- pha_comp_change |> 
  group_by(year, component) |> 
  summarise(value = sum(value)) |> 
  mutate(year = as.numeric(year),
         percent = value/sum(value)) |> 
  filter(year >= 2016 & year != 2020) |> 
  mutate(year = as.character(year))

ggplot(comp_chg_summary,
       aes(x=year,
           y=percent,
           fill=component)) + 
  geom_bar(stat="identity", position = "stack") +
  labs(title="Components of population change",
       subtitle = "Share of regional population growth by source",
       caption = "Note: Data not available for 2020.\nSource: U.S. Census Bureau, Population Estimates Program.") +
  theme(axis.title = element_blank(),
        panel.background = element_blank(),
        legend.title = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major.y = element_line(color = "grey95",
                                          size = 0.05)) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = viridis(3))

```


## Young adults are shacking up.

<br>

```{r}
#| label: hh-adultchild-plot
#| echo: false

b09021_data <- read_rds("../data/b09021_data.rds") |> 
  filter(NAME %in% pha_names) |>
  group_by(arrangement) |> 
  summarise(`2016` = sum(`2016`),
            `2020` = sum(`2020`)) |> 
  mutate(pct_change = (`2020`-`2016`)/`2016`)

ggplot(b09021_data,
       aes(x = reorder(arrangement, pct_change),
           y = pct_change,
           fill = arrangement,
           color = arrangement,
           label = label_percent()(pct_change))) +
  geom_col(position = "dodge") +
  geom_text(aes(y = pct_change + sign(pct_change)/75,
                label = label_percent(accuracy = 0.1)(pct_change)),
            size = 4) +
  scale_y_continuous(labels = label_percent(),
                     breaks = c(-0.075,-0.05, -0.025, 0, 0.025, 0.05, 0.075, 0.1, 0.125, 0.15, 0.175),
                     limits = c(-0.08,0.18)) + 
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
  labs(title = "Change in 18-34 year old population by living arrangement",
       subtitle="2016 to 2020",
       caption = "Source: U.S. Census Bureau, American Community Survey, Table B09021.") +
  theme(axis.title = element_blank(),
        panel.background = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "none",
        panel.grid.major.y = element_line(color = "grey95",
                                          size = 0.05),
        axis.ticks = element_blank())

```

## Lowest-payed workers saw the biggest (relative) pay bumps.

```{r}
#| label: wage-pct-plot
#| echo: false

oews_pct <- read_rds("../data/oews_pct.rds")

ggplot(oews_pct,
       aes(x = pct_change,
           y = wage,
           fill = wage)) +
  geom_col() +
  geom_text(aes(label = label_percent(accuracy = 0.1)(pct_change),
                color = wage),
            nudge_x = 0.01) +
  labs(title = "Percent change in annual wage by wage percentile",
       subtitle = "Richmond, VA MSA | May 2019 to May 2021",
       caption = "Source: U.S. Bureau of Labor Statistics, Occupational Employment and Wage Statistics.") +
  scale_x_continuous(labels = label_percent()) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank(),
        panel.grid.major.y = element_blank(),
        legend.position = "none",
        panel.grid.major.x = element_line(color = "grey95",
                                          size = 0.05))

```

# Housing supply and market changes

## Racial homeownership gaps remain the status quo.

```{r}
#| label: ho-race-plot
#| echo: false

b25003_race_data <- read_rds("../data/b25003_race_data.rds") |> 
  filter(locality %in% pha_names) |> 
  group_by(year, tenure, race) |> 
  summarise(estimate = sum(estimate)) |> 
  group_by(year, race) |> 
  mutate(pct = estimate/sum(estimate)) |> 
  filter(tenure == "Homeowner")

ggplot(b25003_race_data,
       aes(x = year, 
           y = pct, 
           color = race)) +
  facet_wrap(~fct_reorder(race, pct, .fun = median, .desc = TRUE),
             labeller = label_wrap_gen(10),
             nrow = 1) +
  geom_line(stat = "identity", size = 1) + 
  scale_y_continuous(labels = label_percent(), limits = c(0.05, 0.85), n.breaks = 8) +
  labs(title = "Homeownership rate by race and ethnicity",
       subtitle = "2016 to 2020",
       caption = "Source: U.S. Census Bureau, American Community Survey, Table B25003.") +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none",
        legend.title = element_blank(),
        legend.key = element_blank(),
        panel.background = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(color = "grey95",
                                          size = 0.05),
        axis.text.x = element_text(angle = 90, vjust = 0.5),
        axis.line.x = element_line(color = "grey50",
                                   size = 0.1))

```

## A tight supply is pushing existing home prices closer to new build prices.

\[Median price of new construction and resale\]

## The supply of 2- and 3-bedroom apartments *decreased*.

\[Change in renter-occupied housing units by number of bedrooms\]

## Rising prices for 2- and 3-bedroom apartments reflect higher demand.

\[Average asking rent by bedroom\]

## Housing assistance slide

...

## NOAH slide

...

# Gap analysis

## Affordability slide

...

## Impact slide

...

# Local summaries

## Richmond City

## Chesterfield County

:::: {.columns .nonincremental}

::: {.column width="70%"}

```{r}
#| label: cf-gap
#| echo: false
#| fig-asp: 0.75

gap <- read_rds("../data/tb_18_match.rds") |> 
  filter(county == "Chesterfield County") |> 
  group_by(year, household_income, gapcode) |>
  summarise(estimate = sum(estimate)) |>
  mutate(estimate = case_when(
    gapcode == "Gap" ~ estimate * -1,
    gapcode == "Matches or less than income" ~ estimate
  )) |>
  mutate(gapcode = case_when(
    gapcode == "Gap" ~ "Shortage of units",
    gapcode == "Matches or less than income" ~ "Households at or less than income"
  )) |>
  filter(household_income != "80 percent AMI or greater")

ggplot(gap,
       aes(x = year, 
           y = estimate,
           fill = gapcode)) +
  geom_bar(stat = "identity") +
  facet_grid(~household_income) +
  scale_y_continuous(labels = label_comma()) +
  labs(title = paste0("Chesterfield County rental housing gap by AMI"),
       subtitle = "2015 to 2018",
       caption = "Source: U.S. Department of Housing and Urban Development,\nComprehensive Housing Affordability Strategy (CHAS), Table 7.") +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank(),
        panel.grid.major.y = element_line(color = "grey95",
                                          size = 0.05),
        legend.position = "bottom",
        legend.title = element_blank())

```

:::

::: {.column width="30%"}

-   In 2018, there was a shortage of 7,569 rental units for households below 80% AMI

-   Greatest need in the county has been among households below 50% AMI

-   But the need for rental at the higher end has been increasing (15% increase)

:::

::::

## Hanover County

## Henrico County

## Next steps

-   Feedback from PHA Board and staff

-   Complete final updates on report (incorporating edits and suggestions)

-   Consistent branding and visualizations

-   Provide updates to data based on timing

# Discussion
